RBE/CS549 Project: Boat Tracking/Detection

### 2017-12-15 paper/main.tex
Minor changes and punctuation

We have approximately 240GB of HD video from a GoPro
mounted on the prow of a large motorboat. While we
do not have the processing power to fully take advantage
of every frame of data, we will pull reasonable samples
of data to work with.  Specifically, we will generate a test dataset of reduced sized that includes the following important characteristics:

@@

\section{Pixel Level Water Segmentation}
In the first part of the pipeline, we look at individual images and attempt to
provide a pixel level detection of potential objects to track.  This step of the
pipeline provides a confidence value for each pixel in the frame for confidence
that the pixels is an object to track.
The image is converted to HSV color space from the original RGB colorspace since
this gives us better segmentation results. The HSV space has better color
properties for segmentation since the water is typical near one particular
color, but can have variations in brightness.  This also allows the segmentation
to easily segment non-water pixels by the hue value, since the water tends to be
near the blue and green hue values in our test images.

\subsection{Na\"ive K-Means Classifier}
We initially started with a basic two-class k-means classifier in order to
observe how well it would perform.  The classifier was trained on the first
image in the video sequence, and the L2 distance to the two cluster centers was
used to classify each pixel in subsequent frames.  Surprisingly enough, the
2-class k-means classifier worked relatively robustly on the videos that we
provided the algorithm, as show in figure \ref{fig:kmeans}.  The water pixels
are mostly labeled as one class, and pixels that not water are mostly labeled as
another class with a little bit of misclassification noise.  The boat and it's
wake are clearly labeled as potential obstacles to track, which is what we want.
The land areas are labeled as water pixels, but for our implementation we want
to ignore tracking the land anyways. There is a slight bit of noise in the water
reflections towards the right side of figure \ref{fig:kmeans}, but this na\"ive
method still provides very useful information.

K-means does not always associate the same class identifier with the water and
non-water pixels each time, so we must perform an intermediary step to detect
which class identifier it associates with water pixels.  To do so, we find the
number of pixels in each class in the bottom half of the image, and use the sums
to vote for which class is associated with water pixels. This assumes that the
boat is upright, and that a majority of the pixels in the bottom half of the
image are water, and holds true for all of the cases we observed.  If the water
class ID is not what was expected, the cluster centers are swapped so that the
water class ID is always a known value.

\begin{figure}
\includegraphics[width=7.8cm]{hsv_kmeans2_orig}
\includegraphics[width=8.5cm]{hsv_kmeans2_result}
\centering
\caption{Na\"ive K-Means Classifier Results}
\label{fig:kmeans}
\end{figure}

\subsection{Wave Reflection Suppression}
Although there is some noise in the classification results, the k-means
classifier does a decent job of selecting pixels that are areas of water. In
order to suppress false detections as seen in the right side of the classifier
results in figure \ref{fig:kmeans}, we look at the distance values between the
pixel's HSV value and the water pixel cluster center to estimate how similar the
pixel is to the water class.  The wave reflections observed in figure
\ref{fig:kmeans} tend to very close to the cluster center while true obstacle
pixels tend to be further away from the cluster center.  By using this measure
as a confidence value, we can suppress regions that are more likely to be water,
and emphasize regions that are more likely to be objects to track.  Using the
distance metric directly provided decent results, but squaring the distance
metric value provided significantly better suppression of water and emphasis of
potential objects.  The squared value was then used as a direct measure of
confidence that each pixel was a potential object to track.

\subsection{Sky Suppression}
The k-means classifier provides information about which pixels are water and
which pixels are not water, but we would also like the algorithm to only focus
on potential obstacles in the water.  The region of the sky shown in figure
\ref{fig:kmeans} is classified as a non-water pixel, and we would like to
suppress this response to avoid detecting the sky as an obstacle. With only the
wave reflection suppression confidences, most of the sky is suppressed, but
tends to have higher confidences near land as shown in figure
\ref{fig:skysuppress}.  Since the sky pixels are all connected, and define a
very large region, we used the Matlab \texttt{bwareaopen} function that removes binary
pixels in connected areas that are smaller than a particular size.  By looking
only at pixels that are classified as "not water", we can detect the sky region
by filtering for pixel clusters that are larger than 1/4 of the area of the
image, and set the confidence values for those pixels to zero to suppress them.
In some cases, potential objects to track were touching the sky region and the
\texttt{bwareaopen} results tended to remove those object to track.  To fix this, an
erode operation is done on the non-water class pixels to better separate the sky
and potential objects to track.

\begin{figure}
\includegraphics[height=5cm]{hsv_confidence}
\includegraphics[height=5cm]{hsv_kmeans2_suppressed}
\centering
\caption{Sky Suppression Before and After}
\label{fig:skysuppress}
\end{figure}

@@

For the underlying optical flow algorithm, we use Farneback's
\cite{farneback2003} which is accessible in matlab as
\texttt{opticalFlowFarneback}.

@@

In practice, we just used the default options to the matlab method (which uses
a 3-level pyramid scaling image size by a factor of 2, with a 5-pixel
neighborhood size and a Gaussian filter on the output of size 15x15), as these
options tended to produce good results.

@@

In Figure \ref{fig:example_flow} we show the output of
\texttt{opticalFlowFarneback} on a given pair of frames. The most immediate need
in this image is to try and account for the field of flow produced along the
water. This corresponds to the real movement of the plane of the water relative
to the boat. We can attempt to model this with the following assumptions:
\begin{itemize}
\item That the camera is pointed straight and level
\item There is no substantial roll or yaw angular rates of turn occurring
\item The camera is an ideal pinhole camera with no substantial lens distortions
      or the such
\item There is no sideways relative velocity between the boat and the water
\end{itemize}

@@

To actually perform the least-squares fitting we use \texttt{fitlm} in
matlab. Normally, this just uses a normal least-squares estimation
(similar to doing \texttt{A \ b}), but adds some additional
options. We tried to make use of these options, including a outlier
rejection algorithm, but the effects were minimal and substantially
increased the processing time required to execute the code.

@@

When we compare the expected magnitude of the optical flow to the
actual magnitude of the optical flow in Figure \ref{fig:diff_magnitude}
(this is the same image as Figure \ref{fig:example_flow}),
in which you can see a few features:
\begin{itemize}
\item The boat in the left half of the image shows up as a cyan blob, distinct
      from the water around it.
\item The curved metal in the bottom left screws with the optical flow
      algorithm, causing substantial variation from the model.
\item In the bottom right, due to either lens distortion or hard to measure
      waves, the optical flow varies substantially from the model.
\end{itemize}

As a result of the poor results in the very bottom of the image and in the
area around the metal bar, we completely ignore and filter out these regions,
as they are relatively unlikely to contain obstacles (or if they did, our
boat would be running them over anyways).

We also note that in general objects appearing more distantly in the center of
the image will be hard to identify, as the absolute value of the difference
in magnitude from the modelled magnitude will be low (because we are looking
at optical flow in the image, not real life, so far away objects have
low velocities in the image), and so simple thresholds will not work well.
We experimented with reducing the required threshold in the center of the image,
which did improve tracking.

@@

\begin{enumerate}
\item For each pair of new objects, as well as each pairing of one new and one
      old object, we compute a difference score that is the sum of:
  \begin{itemize}
  \item The difference in velocity direction
  \item One minus the quotient in velocity magnitudes (speed), such that the
        larger of the two speeds shall always be in the denominator
  \item The difference in the positions (L2-norm) of the two objects, scaled
        along each axis by the mean width and mean height between the two
        objects (such that if the objects are perfectly aligned in one dimension
        and touching edges in the other, they will be a distance of 1.0 apart).
  \end{itemize}
\item We threshold every difference, leaving a matrix of object ``neighbors''.
\item Iterating over the newly created objects (in arbitrary order), we
      combine that object with it's immediate neighbors, and then remove
      all the objects that were combined from future
      consideration\footnote{Ideally, we might do some sort of search where we
      combine neighbors-of-neighbors and so on, but for simplicities' sake, and
      to avoid situations where we end up combining large numbers of objects
      that all are tenuously connected, we just explore one step out from any
      given object.}.
\item In order to combine each set of objects that we associated, we:
  \begin{itemize}
    \item Create a bounding box that surrounds all of the combined objects
    \item Give the amalgamation the label (and time-of-existence) of the oldest
          object being included, on the assumption that we are interested
          in keeping tracking going across the longest period of time possible.
    \item Perform a weighted average of the velocities, weighting by object area.
    \item Take the maximum of the confidence scores for the individual objects,
          assign it to the result object, and then increment it up slightly on
          the assumption that any time we are combining more than one object,
          we are probably becoming more certain in the overall object's
          existence.
  \end{itemize}
\item Any old objects that were not associated with blobs in the current frame
      have their confidence scores reduced and, if they dip below a confidence
      of 0, are deleted permanently.
\item Finally, for every object being tracked, we keep a history of its last 20
      positions. Once an object accumulates a history of more than around 5
      frames (to reduce potential for individual frames causing too much noise),
      we start to compare the object's overall average position change to its
      velocity. If the two differ by more than a particular threshold, then we
      reduce the confidence score again. This helps to mitigate object resulting
      from the wake of the boat, as such objects generally have optical flows
      inconsistent with where the object actually moves frame-to-frame.
\end{enumerate}

@@

As we are interested in ultimately implementing some of these algorithms on our
actual robot, we are interested in the run time of our algorithms. These
are extremely rough numbers, from running only single-threaded on a laptop CPU,
but were reasonably consistent across runs:
\begin{itemize}
\item Around 3 seconds total per frame
\item Around 1 second to call \texttt{opticalFlowFarneback} on the full
      1920x1080 image
\item Around half a second to perform the linear fit described in section
      \ref{expected_flow}
\item Around a second and a half spread out among other steps
\end{itemize}

@@

While the algorithms do work reasonably well for footage from these particular
conditions, we had troubles processing some parts of the data. In particular,
early in certain mornings when not only was the sun interfering with the camera,
but also in situations such as Figure \ref{fig:mirror}, in which the water is
relatively flat (reducing texture) and highly reflective, meaning that it is no
longer particularly blue.

@@

For future work, we need to actually improve the timing (as mentioned above) to
work in real time, improve robustness for a greater variety of conditions, and
to increase the generality to function on a less stable platform--our boat will
certainly be pitching and rolling substantially more than the platform we had
for this footage.

>>>>

We have approximately 240GB of HD video from a GoPro
mounted on the prow of a large motorboat. While we
do not have the processing power to fully take advantage
of every frame of data, we pulled reasonable samples
of data to work with.  Specifically, we generated a test dataset of reduced sized that includes the following important characteristics:

@@

\section{Pixel Level Water Segmentation}
In the first part of the pipeline, we look at individual images and attempt to
provide a pixel level detection of potential objects to track.  This step of the
pipeline provides a confidence value for each pixel in the frame for confidence
that the pixels is an object to track.
The image is converted to HSV color space from the original RGB color space since
this gives us better segmentation results. The HSV space has better color
properties for segmentation since the water is typical near one particular
color, but can have variations in brightness.  This also allows the segmentation
to easily segment non-water pixels by the hue value, since the water tends to be
near the blue and green hue values in our test images.

\subsection{Na\"ive K-Means Classifier}
We initially started with a basic two-class k-means classifier in order to
observe how well it would perform.  The classifier was trained on the first
image in the video sequence, and the L2 distance to the two cluster centers was
used to classify each pixel in subsequent frames.  Surprisingly enough, the
2-class k-means classifier worked relatively robustly on the videos that we
provided the algorithm, as show in figure \ref{fig:kmeans}.  The water pixels
are mostly labeled correctly, and pixels that are not water are mostly labeled correctly with a little bit of misclassification noise.  The boat and it's
wake are clearly labeled as potential obstacles to track, which is what we want.
The land areas are labeled as water pixels, but for our implementation we want
to ignore tracking the land anyways. There is a slight bit of noise in the water
reflections towards the right side of figure \ref{fig:kmeans}, but this na\"ive
method still provides very useful information.

K-means does not always associate the same class identifier with the water and
non-water pixels each time, so we must perform an intermediary step to detect
which class identifier it associates with water pixels.  To do so, we find the
number of pixels in each class in the bottom half of the image, and use the sums
to vote for which class is associated with water pixels. This assumes that the
boat is upright, and that a majority of the pixels in the bottom half of the
image are water, which holds true for all of the cases we observed.  If the water
class ID is not what was expected, the cluster centers are swapped so that the
water class ID is always a known value.

\begin{figure}[H]
\includegraphics[width=7.8cm]{hsv_kmeans2_orig}
\includegraphics[width=8.5cm]{hsv_kmeans2_result}
\centering
\caption{Na\"ive K-Means Classifier Results}
\label{fig:kmeans}
\end{figure}

\subsection{Wave Reflection Suppression}
Although there is some noise in the classification results, the k-means
classifier does a decent job of selecting pixels that are areas of water. In
order to suppress false detections as seen in the right side of the classifier
results in figure \ref{fig:kmeans}, we look at the distance values between the
pixel's HSV value and the water pixel cluster center to estimate how similar the
pixel is to the water class.  The wave reflections observed in figure
\ref{fig:kmeans} tend to be very close to the cluster center while true obstacle
pixels tend to be further away from the cluster center.  By using this measure
as a confidence value, we can suppress regions that are more likely to be water,
and emphasize regions that are more likely to be objects to track.  Using the
distance metric directly provided decent results, but squaring the distance
metric value provided significantly better suppression of water and emphasis of
potential objects.  The squared value was then used as a direct measure of
confidence that each pixel was a potential object to track.

\subsection{Sky Suppression}
The k-means classifier provides information about which pixels are water and
which pixels are not water, but we would also like the algorithm to only focus
on potential obstacles in the water.  The region of the sky shown in figure
\ref{fig:kmeans} is classified as a non-water pixels, and we would like to
suppress this response to avoid detecting the sky as an obstacle. With only the
wave reflection suppression confidences, most of the sky is suppressed, but
tends to have higher confidences near land as shown in figure
\ref{fig:skysuppress}.  Since the sky pixels are all connected, and define a
very large region, we used the Matlab \texttt{bwareaopen} function that removes binary
pixels in connected areas that are smaller than a particular size.  By looking
only at pixels that are classified as "not water", we can detect the sky region
by filtering for pixel clusters that are larger than 1/4 of the area of the
image, and set the confidence values for those pixels to zero to suppress them.
In some cases, potential objects to track were touching the sky region and the
\texttt{bwareaopen} results tended to remove those object to track.  To fix this, an
erode operation is performed on the non-water class pixels to better separate the sky
and potential objects to track.

\begin{figure}[H]
\includegraphics[height=5cm]{hsv_confidence}
\includegraphics[height=5cm]{hsv_kmeans2_suppressed}
\centering
\caption{Sky Suppression Before and After}
\label{fig:skysuppress}
\end{figure}

@@

For the underlying optical flow algorithm, we use Farneback's
\cite{farneback2003} which is accessible in Matlab as
\texttt{opticalFlowFarneback}.

@@

In practice, we just used the default options to the Matlab method (which uses
a 3-level pyramid scaling image size by a factor of 2, with a 5-pixel
neighborhood size and a Gaussian filter on the output of size 15x15), as these
options tended to produce good results.

@@

In Figure \ref{fig:example_flow} we show the output of
\texttt{opticalFlowFarneback} on a given pair of frames. The most immediate need
in this image is to try and account for the field of flow produced along the
water. This corresponds to the real movement of the plane of the water relative
to the boat. We can attempt to model this with the following assumptions:
\begin{itemize}
\item That the camera is pointed straight and level
\item There is no substantial roll or yaw angular rates of turn occurring
\item The camera is an ideal pinhole camera with no substantial lens distortions
\item There is no sideways relative velocity between the boat and the water
\end{itemize}

@@

To actually perform the least-squares fitting we use \texttt{fitlm} in
Matlab. Normally, this just uses a normal least-squares estimation
(similar to doing \texttt{A \ b}), but adds some additional
options. We tried to make use of these options, including a outlier
rejection algorithm, but the effects were minimal and substantially
increased the processing time required to execute the code.

@@

When we compare the expected magnitude of the optical flow to the
actual magnitude of the optical flow in Figure \ref{fig:diff_magnitude}
(this is the same image as Figure \ref{fig:example_flow}),
in which you can see a few features:
\begin{itemize}
\item The boat in the left half of the image shows up as a cyan blob, distinct
      from the water around it.
\item The curved metal in the bottom left affects the optical flow
      algorithm, causing substantial variation from the model.
\item In the bottom right, due to either lens distortion or hard to measure
      waves, the optical flow varies substantially from the model.
\end{itemize}

Due to the poor results in the very bottom of the image and in the
area around the metal bar, we completely ignore and filter out these regions,
as they are relatively unlikely to contain obstacles (or if they did, our
boat would be running them over anyways).

We also note that, in general, objects appearing more distantly in the center of
the image will be hard to identify, as the absolute value of the difference
in magnitude from the modelled magnitude will be low (because we are looking
at optical flow in the image, not real life, so far away objects have
low velocities in the image), and so simple thresholds will not work well.
We experimented with reducing the required threshold in the center of the image,
which did improve tracking.

@@

\begin{enumerate}
\item For each pair of new objects, as well as each pairing of one new and one
      old object, we compute a difference score that is the sum of:
  \begin{itemize}
  \item The difference in velocity direction
  \item One minus the quotient in velocity magnitudes (speed), such that the
        larger of the two speeds shall always be in the denominator
  \item The difference in the positions (L2-norm) of the two objects, scaled
        along each axis by the mean width and mean height between the two
        objects (such that if the objects are perfectly aligned in one dimension
        and touching edges in the other, they will be a distance of 1.0 apart).
  \end{itemize}
\item We threshold every difference, leaving a matrix of object ``neighbors''.
\item Iterating over the newly created objects (in arbitrary order), we
      combine that object with it's immediate neighbors, and then remove
      all the objects that were combined from future
      consideration\footnote{Ideally, we might do some sort of search where we
      combine neighbors-of-neighbors and so on, but for simplicities' sake, and
      to avoid situations where we end up combining large numbers of objects
      that all are tenuously connected, we just explore one step out from any
      given object.}.
\item In order to combine each set of objects that we associated, we:
  \begin{itemize}
    \item Create a bounding box that surrounds all of the combined objects
    \item Give the amalgamation the label (and time-of-existence) of the oldest
          object being included, on the assumption that we are interested
          in maintaining tracking going across the longest period of time possible.
    \item Perform a weighted average of the velocities, weighting by object area.
    \item Take the maximum of the confidence scores for the individual objects,
          assign it to the result object, and then increment it up slightly on
          the assumption that any time we are combining more than one object,
          we are probably becoming more certain in the overall object's
          existence.
  \end{itemize}
\item Any old objects that were not associated with blobs in the current frame
      have their confidence scores reduced and, if they dip below a confidence
      of 0, are deleted permanently.
\item Finally, for every object being tracked, we keep a history of its last 20
      positions. Once an object accumulates a history of more than around 5
      frames (to reduce potential for individual frames causing too much noise),
      we start to compare the object's overall average position change to its
      velocity. If the two differ by more than a particular threshold, then we
      reduce the confidence score again. This helps to mitigate object resulting
      from the wake of the boat, as such objects generally have optical flows
      inconsistent with where the object actually moves frame-to-frame.
\end{enumerate}

@@

As we are interested in ultimately implementing some of these algorithms on our
actual robot, we are interested in the run time of our algorithms. These
are extremely rough numbers, from running only single-threaded on a laptop CPU,
but were reasonably consistent across runs:
\begin{itemize}
\item Around 3 seconds total per frame
\item Around 200 msec for the HSV segmentation
\item Around 1 second to call \texttt{opticalFlowFarneback} on the full
      1920x1080 image
\item Around half a second to perform the linear fit described in section
      \ref{expected_flow}
\item Around a second and a half spread out among other steps
\end{itemize}

@@

While the algorithms do work reasonably well for footage from these particular
conditions, we had troubles processing some parts of the data. In particular,
early mornings when not only was the sun interfering with the camera,
but also in situations such as Figure \ref{fig:mirror}, in which the water is
relatively flat (reducing texture) and highly reflective, meaning that it is no
longer particularly blue. The algorithm also struggles when pointed in the
 direction of the sun, as the reflections off of the water tend to oversaturate the image sensor.

@@

For future work, we need to actually improve the timing (as mentioned above) to
work in real time, improve robustness for a greater variety of conditions, and
to increase the generality to function on a less stable platform--our boat will
certainly be pitching and rolling substantially more than the platform we had
for this footage, and the weather conditions may be less ideal.
